# The Soaring Cost of Storms: Severe Weather Effects on Health and Property

## Synopsis

In this report we describe the most damaging storm event types to human health and property from 1950 to recent years. Property damage due to storm events has risen dramatically in the past fifeteen years. To demonstrate this claim we downloaded the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database for the years 1950 to 2011. We list the most damaging storm types causing human casualties (deaths and injuries). We also list the most damaging storm types causing property and crop losses. We found that tornados and floods caused the most damage respectively. We also found that since 1995, the dollar cost of storms has soared.

## Data Processing

We downloaded the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database from [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) on 2015-01-10. It contains storm data from 1950 to 2011. We load the raw data from the .csv.bz2 file this way:

```{r, echo=TRUE, cache=TRUE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(ggplot2))

if (!exists('storm')) {
    storm <- read.csv('./repdata_data_StormData.csv.bz2')    
}
```

### Transformations

#### Dates
We transform the beginning date (BGN_DATE) field so we have a proper date format to work with. We create some convenience fields for year and five year period to make it easy to analyse storm data over different aggregated time horizons. The five year periods are the years 00 - 04 and 05 - 09 for each decade and named after the first year in the period. For example, the 2005 period contains the years 2005 to 2009 inclusive.

```{r}
# split out date from BGN_DATE's date/time factor and cast as Date
storm$evdate <-sapply(as.character(storm$BGN_DATE), function(x){ strsplit(x, split=c(' '))[[1]][1] })
storm$evdate <- as.Date(storm$evdate, format="%m/%d/%Y")

# Create separate year and fiveyear period variables 
storm$year <- lubridate::year(storm$evdate)
storm$fiveyear = paste(substr(storm$year, 0,3), ifelse(substr(storm$year, 4, 4) < 5, 0, 5), sep='')
```

#### Population Impact

Sum fatalities and injuries as 'casualties' to simplify analysis of the health effects of storms.
```{r}
storm$casualties <- storm$FATALITIES + storm$INJURIES
```

#### Property Damage

It is likely that the PROPDMG and CROPDMG fields are intended to be modified according to the values found in PROPDMGEXP and CROPDMGEXP. However, since we find no description in the official sources as to how this should work, this present analysis assumes the following: if the PROPDMGEXP and CROPDMGEXP are equal to 'm' or 'M' the damages is in millions of dollars. All other values of PROPDMGEXP and CROPDMGEXP are assumed to signify thousands of dollars (""  "-" "?" "+" "0" "1" "2" "3" "4" "5" "6" "7" "8" "B" "H" "h" "K"). (cf. [Storm Events Database](http://www.ncdc.noaa.gov/stormevents/), [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf), [National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf) )

Sum property damage and crop damage as 'damages' to simplify analysis of the health effects of storms.
```{r}
# Normalize the PROPDMGEXP and CROPDMGEXP fields
storm$PROPDMGEXP <- ifelse(storm$PROPDMGEXP == 'm' | storm$PROPDMGEXP == 'M', 'm', 'k')
storm$CROPDMGEXP <- ifelse(storm$CROPDMGEXP == 'm' | storm$CROPDMGEXP == 'M', 'm', 'k')

# Calculate true values
storm$PROPDMG <- ifelse(storm$PROPDMGEXP == 'm', storm$PROPDMG * 1000000, storm$PROPDMG * 1000)
storm$CROPDMG <- ifelse(storm$CROPDMGEXP == 'm', storm$CROPDMG * 1000000, storm$CROPDMG * 1000)

# Combined damages
storm$damages <- storm$PROPDMG + storm$CROPDMG
```

### Shape of the data

Our main focus will be on event type, dates, property damage, and casualties.
```{r}
storm[sample(1:900000, 5), c(2, 8, 23:27, 38:42)]
```
__*Table 1. Random sample from the data set*__

#### Event type

The EVTYPE field records the type of storm event. It is far from a well-curated taxonomy. We acknowledge this problem with the data, but choose not to address it in this analysis

```{r}
sort(grep('Wind', unique(storm$EVTYPE), value = TRUE))
```
__*Table 2. Wind event types showing many overlapping codes*__

```{r}
sort(grep('Snow', unique(storm$EVTYPE), value = TRUE))
```
__*Table 3. Snow event types showing many overlapping codes*__

## Results

### Long term - 1950 - 2011

To determine the storm events most harmful to population health, we can tapply over EVTYPE summing casualties, sort, and display the top ten:

```{r}
all.casualties.nums <- with(storm, tapply(casualties, EVTYPE, sum, na.rm = TRUE))
all.casualties <- data.frame(casualties = all.casualties.nums, event <- names(all.casualties.nums))
all.casualties$casualties <- as.numeric(all.casualties$casualties)
all.casualties <- dplyr::arrange(all.casualties, desc(casualties))
names(all.casualties) <- c('casualties', 'event')
head(all.casualties, 10)
```
__*Table 4. Storm events most harmful to population health, 1950 - 2011*__

Tornados have been far and away the leading cause of death and injury from storm events since 1950.

The same technique - using damages instead of casualties - yields the storm events that have had the greatest economic consequences (property and crops).

```{r}
all.damage.nums <- with(storm[storm$fiveyear == '2005',], tapply(damages, EVTYPE, sum, na.rm = TRUE))
all.damage <- data.frame(property.damage = all.damage.nums, event <- names(all.damage.nums))
all.damage$property.damage <- as.numeric(all.damage$property.damage)
all.damage <- dplyr::arrange(all.damage, desc(property.damage))
names(all.damage) <- c('property.damage', 'event')
head(all.damage, 10)
```
__*Table 5. Storm events with the greatest economic consequences, 1950 - 2011*__

### Rising cost of storms

There has been a dramatic rise in the property damage caused by storms.

```{r}
fiveyear.damage.nums <- tapply(storm$damages, storm$fiveyear, sum)
fiveyear.damage <- data.frame(damages = fiveyear.damage.nums[1:12] / 1000000, fiveyear = names(fiveyear.damage.nums[1:12]))
fiveyear.plot <- ggplot2::ggplot(data = fiveyear.damage, ggplot2::aes(x=fiveyear, y=damages, width=0.3)) +
    ggplot2::geom_bar(stat='identity') +
    ggplot2::ggtitle("Storm Damage By Five Year Period - 1950-2009") +
    ggplot2::xlab('Data source: U.S. National Oceanic and Atmospheric Administration (NOAA)') +
    ggplot2::ylab('Property and crop damage (millions of U.S. dollars)') 
print(fiveyear.plot)
```

__*Fig 1. Storm Damage By Five Year Period - 1950-2009*__
